---
title: "Examples For Lecture 8"
author: "JMG"
date: "9/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggformula)
```

## Examples for Point Estimation 

### Example 5.4

In this example we are given that the sampling distribution for the sample proportion $\hat{p}$ is $N(\mu_{\hat{p}}=0.88,SE_{\hat{p}}=0.01)$. 

For an $\hat{p}$ to be within 2% of the population proportion $p=0.88$, we must have $\hat{P}$ between $0.88-0.02=0.86$ and $0.88+0.02=0.9$. The corresponding $Z$-scores are

$$Z_{0.86}=\frac{0.86-0.88}{0.01}=-2, \ \text{ and } \ Z_{0.9}=\frac{0.9-0.88}{0.01}=2$$

We need to compute the area under the standard normal density function between -2 and 2. That is, the area shown below:
```{r}
gf_dist("norm",fill = ~(x >= -2 & x<= 2),geom="area")
```

We can compute this area as
```{r}
pnorm(2) - pnorm(-2)
```

Thus, 95.45% of the time, we expect the sample proportion to be within 2% of the population proportion $p=0.88$. 

### Exercise 5.1


(a) mean


(b) mean


(c) proportion 


(d) mean 


(e) proportion 


### Exercise 5.3

(a) population is those chips manufactured during the week the sample is taken



(b) proportion of defective chips



(c) observed $\hat{p} = \frac{27}{212}\approx 0.13$



(d) standard error for the sample proportion




(e) $SE_{\hat{p}} \approx  \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}=\sqrt{\frac{0.13(1-0.13)}{212}} \approx 0.023$



(f) We expect that 95% of the time the observed sample proportion will fall within 1.96 times the standard error. That is about within
```{r}
(0.13 - 1.96* 0.023)
(0.13 + 1.96* 0.023)
```

Since 10% is well within this interval, we should not be surprised.



(g) 

```{r}
sqrt((0.1*(1-0.1))/212)
```

There isn't much of a difference. 


## Examples for Confidence Intervals

### Guided Practice 5.6

If we want to be very certain we capture the population proportion, should we use a wider interval or a smaller interval? 


### Example 5.8

We are given that $\hat{p}$ follows a normal distribution and has a standard error $SE_{\hat{p}}=0.01$. Then a 95% confidence interval (CI) is
```{r}
0.887 + c(-1,1)*1.96*0.01
```

This tells us that if we were to repeatedly sample from the same population many many times, we would expect the true proportion to fall within the 86.7% to 90.7%  range 95% of the time. 

### Guided Practice 5.9

The following plot shows the area under the standard normal density curve that lies within 2.58 standard deviations of the mean. 

```{r}
gf_dist("norm",fill = ~(x >= -2.58 & x<= 2.58),geom="area")
```

We can compute this area by
```{r}
pnorm(2.58) - pnorm(-2.58)
```

Thus, about 99% of the time, a normal random variable will fall within 2.58 standard deviations from the mean. 

### Exercise 5.8

```{r}
(z_ast <- qnorm(0.005)) 
(CI <- 0.52 + c(1,-1)*z_ast*0.024)
```


### Guided Practice 5.27

How could we reduce the Type 2 Error rate in US courts? What influence would this have on the Type 1 Error rate? 

To lower the Type 2 Error rate we need to convict more guilty people. This will result in more wrongful convictions, raising the Type 1 Error rate. 
