---
title: "Lecture 7"
subtitle: "The Normal Distribution"
author: "JMG"
institute: "MATH 204"
date: "Tuesday, September 21"
output:
  xaringan::moon_reader:
    css: [rladies, rladies-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(openintro)
library(tidyverse)
library(ggformula)
library(patchwork)
library(latex2exp)
library(kableExtra)
```


# Learning Objectives

In this lecture, we will

- Study the **normal distribution** and learn how to compute probability values for a random variable that follows a normal distribution.

--

- The normal distribution is a continuous distribution. We will discuss the so-called normal curve which is the probability density function for a normal distribution. 

--

- We will also learn about the expected value and variance for a normal random variable. 

--

- This lecture corresponds to section 4.1 in the textbook, and you are encouraged to watch the lecture video included in the next slide. 


---

# Normal Distribution Video

```{r,echo=FALSE}
vembedr::embed_url("https://youtu.be/S_p5D-YXLS4") %>%
  vembedr::use_align("center")
```

---

# Samples from a Normal Distribution

- The following histogram shows 5,000 random samples from a normal random variable:

```{r norm_samps,fig.height=5}
tibble(x=rnorm(5000,66,3.5)) %>% gf_histogram(~x,color="black")
```

---

# Normal Histogram Discussion 

- Think of the histogram on the previous slide as showing sample data for measurements of human heights in inches.  

--

- What are the key features of the histogram?

--

- It is unimodal, highly symmetric, and centered at the mean.

--

- Do you think that such data could reasonably correspond to measurements of human heights in inches?

--

- Do you think it is reasonable to treat measurements of human heights in inches as a continuous variable? 

---

# Normal Distribution Facts

> Many random variables are nearly normal, but none are exactly normal. Thus the normal distribution, while not perfect for any single problem, is very useful for a variety of problems. We will use it in data exploration and to solve important problems in statistics. 

--

- Let's spend some time to develop some intuition for how the normal distribution is often used in practice.

--

- We start by generating some data that is not necessarily normally distributed. 

---

# Sampling a Mean

- We are going to play a game. We proceed as follows: 

--

  1) Sample 15 values from a binomial random variable with $n=25$ and $p=0.5$. You can think of this as doing 15 rounds of an experiment where each time we flip a fair coin 25 times and count the number of heads that we obtain.
  
--

  2) We compute and record the mean of the 15 values obtained in step 1. 

--

  3) We repeat steps 1 & 2 a very large number of times, say 2,500. 
  
--

- Here is the first few rows of a data frame that contains the data we acquire by playing our game.

```{r samp_mean,echo=FALSE}
get_means <- function(i){
  return(mean(rbinom(15,25,0.5)))
}

ms <- map_dbl(1:2500,get_means)
tibble(means=ms) %>% head()
```

---

# Plotting Our Data

- Here is a histogram of the data we obtained from the game described on the previous slide:

```{r samp_mean_hist,echo=FALSE,fig.height=4}
tibble(means=ms) %>% gf_histogram(~means,color="black")
```

--

- The mean of our data is `r round(mean(ms),3)`. What is the expected value of a binomial random variable with $n=25$ and $p=0.5$? 

---

# A Normal Density Function

