---
title: "Lecture 13"
subtitle: " "
author: "JMG"
institute: "MATH 204"
output:
  xaringan::moon_reader:
    css: [rladies, rladies-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(openintro)
library(faraway)
library(tidyverse)
library(ggformula)
library(ggmosaic)
library(patchwork)
library(latex2exp)
library(kableExtra)
library(broom)
```


# Intro to Nonparametric Stats

- All of the inferential methods we have studied so far rely on certain distributional assumptions. For example, in many cases we have either assumed or required
data to be normally distributed. 
--

- These distributional assumptions relate to probability distributions, *e.g.*, normal, that depend on parameters, *e.g.*, $\mu$ and $\sigma$ for a normal distribution.

--

- What do we do if the necessary distributional assumptions or requirements can  not be met?


--

- In some cases, there is a [**nonparametric**](https://en.wikipedia.org/wiki/Nonparametric_statistics) parallel to the inferential methods we have studied so far that allow us to weaken the conditions, say for example to conduct a hypothesis test. 

--

- We cover some of these nonparametric statistics in this lecture. Note that this material is not covered in our textbook [OpenIntro Statistics](https://openintro.org/book/os/). 


---

# Parametric Vs. Nonparametric

- Below we list some inferential methods we have already studied together with their parallel nonparametric test.

--

  - t-test for difference of means; [Wilcoxon rank-sum test](https://en.wikipedia.org/wiki/Mann%E2%80%93Whitney_U_test) (or Mann-Whitney-Wilcoxon test)

--

  - paired t-test; [Wilcoxon signed-rank test](https://en.wikipedia.org/wiki/Wilcoxon_signed-rank_test)
  
--

 -  ANOVA (F-test); [Kruskal-Wallis test](https://en.wikipedia.org/wiki/Kruskal%E2%80%93Wallis_one-way_analysis_of_variance). 
 
--

 - We also look at a [sign test](https://en.wikipedia.org/wiki/Sign_test) where we are interested in assessing differences between to groups of paired observations but do not taken into account magnitudes of differences. 


---

# Learning Objectives

- After this lecture, you should

--

  - have a sense of how the rank-sum, signed-rank, Kruskal-Wallis, and sign tests work, 
  
--
 
  - have a sense of when to use the rank-sum, signed-rank, Kruskal-Wallis, and sign tests
  
--

  - know how to conduct these tests using R. 
  
---

# Why Use Nonparametric Stats?

- In general, conclusions drawn from non-parametric methods are not as powerful as the parametric ones.

--

- However, as nonparametric methods make fewer assumptions, they are more flexible, more robust, and easier to run.

--

- Nonparametric methods are especially useful when sample sizes are small. 

--

- [Nonparametric statistics](https://en.wikipedia.org/wiki/Nonparametric_statistics) is a significant and substantial subfield of statistics. However, it should be noted that it is difficult to give a simple and precise definition of nonparametric statistics. 

--

- What we cover in this lecture does not even begin to touch upon many of the most important parts of nonparametric inference. 

---

# Rank-sum Test

- We begin our introduction to nonparametric methods with a discussion of the Wilcoxon rank-sum test which is also known as the Mann–Whitney U test. 

--

- Suppose we have the following question:

> Do math majors get more A's than non-math majors? 

--

- To answer this question, we collect data. For example:
```{r,echo=FALSE}
major <- c(rep("math",10),rep("non",13))
num_as <- c(10,11,9,10,12,15,16,18,19,20,9,6,9,8,7,10,9,8,7,9,10,12,15)
df_1 <- tibble(major=major,num_as=num_as)
df_1 <- df_1[sample(nrow(df_1),replace = FALSE),]
df_1 %>% head()
```
--

- Note that there are 10 math majors and 13 non-math majors:
```{r,echo=FALSE}
table(df_1$major)
```


---

# Plot of Data

- Here is a plot of our data:

```{r,echo=FALSE,fig.height=5}
df_1 %>% ggplot(aes(x=major,y=num_as)) + 
  geom_point()
```

---

# Assumptions

- In the previous data set, there is no reason to expect that the number of A's is normally distributed. Furthermore, the sample sizes are small. 

--

- The Wilcoxon rank-sum test only requires us to assume that

--

  - All of the observations are independent and the responses are at least ordinal in that we can at least say whether of any two observations, which is greater. 
  
--

-  Rank-sum tests

--

  - $H_{0}:$ The distributions of both populations are equal. 
  
--

  - $H_{A}:$ The distributions of the populations are not equal. 
  

---

# Example Rank-Sum Test

- In R, we conduct a rank-sum test using

```{r,warning=FALSE,message=FALSE}
wilcox.test(num_as~major,data=df_1,correct=FALSE)
```

--

- If our significance level is $\alpha=0.05$, then we would reject the null hypothesis that the distribution of the number of A's is the same for the two groups. 

---

# Signed-rank Test

- Now we consider the Wilcoxon signed-rank test. 

--

- Suppose we have the following question:

> Do you work out longer if you listen to music while exercising?

--

- We randomly select 10 students, and record their workout time in minutes for one workout session without music, and for a second workout session with music. The data might look like follows:

```{r}
(no_music <- c(30,25,20,22,36,40,35,27,30,35))
(with_music <- c(32,27,20,25,30,39,30,30,28,32))
```

--

- It is not reasonable to assume that this data is normally distributed. 

---

# Signed-Rank Test

- We want to test the hypothesis:

--

  - $H_{0}:$ There is no difference in the length of workout sessions.
  
--

  - $H_{A}:$ There is a difference in the length of workout sessions. 
  
--

 - The signed-rank test can be used for this test. 
 
---

# Example Signed-Rank Test

- In R, we conduct a signed-rank test using

```{r,warning=FALSE,message=FALSE}
wilcox.test(no_music,with_music,paired=TRUE,correct=FALSE)
```

--

- We fail to reject the null hypothesis at the $\alpha=0.05$ significance level. 

---

# Kruskal-Wallis

- The Kruskal-Wallis test parallels our ANOVA test for comparing many means. 

--

- The null and alternative hypotheses for the Kruskal-Wallis test are

--

  - $H_{0}:$ There is no difference in the distribution of the populations. 
  
--

  - $H_{A}:$ There is a difference in the distributions of the populations.
  
--

- Two conditions for using the Kruskal-Wallis test are 

--

  - each sample must be randomly selected, and 
  
--

  - the size of each sample must be at least 5. 

---

# Example

- Consider again out data for the weights of chicken fed with different types of feed:

```{r,echo=FALSE,message=FALSE,warning=FALSE,fig.height=5}
chickwts %>% ggplot(aes(x=feed,y=weight)) + geom_boxplot()
```

---

# Kruskal-Wallis Test

- Let's run a Kruskal-Wallis test on the chicken weights data:

```{r}
kruskal.test(weight~feed,data=chickwts)
```

--

- If our significance level is $\alpha=0.05$, then we will decide to reject the null hypothesis that the distribution of weights is the same across all groups. 

--

- Let's compare the results from Kruskal-Wallis with results from an ANOVA test. 

---

# Comparing Tests

- ANOVA

```{r}
summary(aov(weight~feed,data=chickwts))
```


--

- Kruskal-Wallis

```{r}
kruskal.test(weight~feed,data=chickwts)
```

---

# Sign Test


