---
title: "Lectures 6 and 7 Worked Examples"
author: "JMG"
date: "9/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggformula)
```

## R Commands for Distributions from Lecture 6

### Examples of Using a Geometric Distribution

1) If $X$ is a geometric random variable with probability of success $p=0.4$, then the probability of getting a success on the 12-th trial is

```{r}
dgeom(11,0.4)
```

Note that this is the same as
```{r}
(1-0.4)^(12-1)*(0.4)
```

2) If $X$ is a geometric random variable with probability of success $p=0.4$, then the probability of getting a success by at least the 5-th trial is
```{r}
pgeom(4,0.4)
```

Note that this is the same as
```{r}
dgeom(0,0.4) + dgeom(1,0.4) + dgeom(2,0.4) + dgeom(3,0.4) + dgeom(4,0.4)
```

which is the same as
```{r}
sum(dgeom(c(0,1,2,3,4),0.4))
```

which is also the same as
```{r}
sum(0.4*(1-0.4)^c(0,1,2,3,4))
```

3) Draw 6 random samples from a geometric distribution with probability of success $p=0.25$
```{r}
rgeom(6,0.25)
```

### Examples Using a Binomial Distribution

1) Let $X$ be a binomial random variable with size $n=12$ and probability of success $p=0.4$. Then the probability of getting 3 successes is
```{r}
dbinom(3,12,0.4)
```

Note that this is the same as
```{r}
choose(12,3)*(1-0.4)^(12-3)*(0.4)^3
```

2) Let $X$ be a binomial random variable with size $n=12$ and probability of success $p=0.4$. Then the probability of getting at most 5 successes is
```{r}
pbinom(5,12,0.4)
```

Note that this is the same as
```{r}
sum(dbinom(c(0,1,2,3,4,5),12,0.4))
```

3) Draw 6 random samples from a binomial distribution with size $n=8$ and probability of success $p=0.25$
```{r}
rbinom(6,8,0.25)
```

## R Commands for Worked Examples from Lecture 7

### Plotting Normal Density Functions

1) Plot a normal density function corresponding to the distribution $N(\mu=12,\sigma=1.5)$.
```{r}
gf_dist("norm",mean=12,sd=1.5)
```

2) Plot the left tail area $x \leq 10$ of a normal density function corresponding to the distribution $N(\mu=12,\sigma=1.5)$
```{r}
gf_dist("norm",mean=12,sd=1.5,fill = ~(x<=10),geom="area")
```

Compute the left tail area that we just plotted.

```{r}
(z <- (10-12)/1.5)

(pnorm(z))

```

3) Plot the right tail area $x \geq 14$ of a normal density function corresponding to the distribution $N(\mu=12,\sigma=1.5)$
```{r}
gf_dist("norm",mean=12,sd=1.5,fill = ~(x>=14),geom="area")
```

Compute the right tail area that we just plotted.

```{r}
(z <- (14-12)/1.5)

(1 - pnorm(z))

```

4) Plot the mid-area $10 \leq x \leq 13$ of a normal density function corresponding to the distribution $N(\mu=12,\sigma=1.5)$
```{r}
gf_dist("norm",mean=12,sd=1.5,fill = ~(x >= 10 & x<= 13),geom="area")
```

Compute the mid-area that we just plotted.

```{r}
(z_a <- (10-12)/1.5)
(z_b <- (13-12)/1.5)

(pnorm(z_b) - pnorm(z_a))

```

### Guided Practice 4.11

```{r}
gf_dist("norm",mean=1100,sd=200,fill = ~(x<=1500),geom="area")
```

Percentile:
```{r}
z <- (1500-1100)/200
pnorm(z)
```

Percentage that did better than Stuart:
```{r}
1 - pnorm(z)
```


### Guided Practice 4.15

(a) 

```{r}
qnorm(0.95,1100,200)
```

(b)

```{r}
qnorm(0.975,1100,200)
```

### Guided Practice 4.20

(a) 1 standard deviation

```{r}
gf_dist("norm",fill = ~(x >= -1 & x<= 1),geom="area")
```


```{r}
pnorm(1) - pnorm(-1)
```

(b) 2 standard deviations

```{r}
gf_dist("norm",fill = ~(x >= -2 & x<= 2),geom="area")
```


```{r}
pnorm(2) - pnorm(-2)
```

(c)

```{r}
gf_dist("norm",fill = ~(x >= -3 & x<= 3),geom="area")
```


```{r}
pnorm(3) - pnorm(-3)
```



### Exercise 4.4

(a)

Men, Ages 30-34 distribution is $N(\mu=4313,\sigma=583)$.

Women, Ages 25-29 distribution is $N(\mu=5261,\sigma=807)$.

(b)

Leo's $Z$-score
```{r}
(4948-4313)/583
```

Mary's $Z$-score
```{r}
(5513-5261)/807
```

(d)

```{r}
gf_dist("norm",mean=4313,sd=583,fill = ~(x<=4948),geom="area")
```



```{r}
z_leo <- (4948-4313)/583
pnorm(z_leo)
```

### Exercise 4.39

(a)

Start by drawing a picture. 

```{r}
(z_score <- qnorm(0.75))
```

(b)

Mean is 1650 and 75-th is 1800

(c)

```{r}
(std_dev <- (1800-1650)/ 0.67) 
```

### Exercise 4.48

(a)


```{r}
dgeom(2,1/4)
```


(b)


```{r}
dbinom(3,5,1/4) + dbinom(4,5,1/4) 
```

(c)

```{r}
dbinom(3,5,1/4) + dbinom(4,5,1/4) + dbinom(5,5,1/4)
```



