---
title: "Lecture 12"
subtitle: " "
author: "JMG"
institute: "MATH 204"
output:
  xaringan::moon_reader:
    css: [rladies, rladies-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(openintro)
library(faraway)
library(tidyverse)
library(ggformula)
library(ggmosaic)
library(patchwork)
library(latex2exp)
library(kableExtra)
library(broom)
```


# Multiple Regression

- Multiple regression builds on the foundations of simple linear regression to allow for more than one predictor. Watch the following video to get started.

```{r,echo=FALSE}
vembedr::embed_url("https://youtu.be/sQpAuyfEYZg") %>%
  vembedr::use_align("center")
```

---

# Learning Objectives

- After this lecture, you should

--

  - know how to fit a multiple regression model using `lm`,
  
--

  - understand and be able to interpret adjusted $R^2$, and 
  
--

  - be able to use diagnostic plots to assess the validity of a linear fit. 
  
---

# Motivating Data

- Consider the `mariokart` data set which consists of auction data from Ebay for the game Mario Kart for the Nintendo Wii. This data was collected in early October 2009.

--

```{r}
head(mariokart)
```

--

- Let's obtain another view of this data.

---

# Glimpse of `mariokart`

```{r}
glimpse(mariokart)
```

--

- **Question:** What features affect the final price (`total_pr`) at which a game is sold? 

---

# A First Model

- As a start, we fit a linear model with the game condition (`cond`) as the only predictor:

--

```{r}
lm_fit <- lm(total_pr~cond,data=mariokart)
summary(lm_fit)
```

---

# Results

- Our first fit predicts that a used game will, on average, go for $6.62 less than a new game will. 

--

```{r,echo=FALSE,fig.height=4,warning=FALSE,message=FALSE}
lm_fit %>% 
  augment() %>% 
  ggplot() + 
  geom_point(aes(x=cond,y=total_pr),alpha=0.5) +
  geom_abline(slope=-6.623,intercept=53.771,color="blue",linewidth=1) + 
  ylim(c(25,80))
```


--

- **Question:** Do you think that the condition of the game alone is sufficient to predict the price of the game? Explain why or why not. 

---

# Adding Predictors

- As we will see, in R it is extremely easy to fit a model with many predictors. WHy might we want to do this? 

--

- We would like to fit a model that includes all potentially important variables simultaneously. 

--


- Multiple regression can help us evaluate the relationship between a predictor variable and the outcome while controlling for the potential influence of other variables. 

--

- Let's fit a more complicated linear model.

---

# Another Fit

```{r}
lm_fit2 <- lm(total_pr~cond+stock_photo+duration+wheels,data=mariokart)
summary(lm_fit2)
```


---

# Results

- Notice that when we have controlled for other features, the condition (new versus used) of the game  has a smaller impact on the price of the game since the slope estimate has gone from -6.6 to -2.6. 