---
title: "Lecture 11"
subtitle: " "
author: "JMG"
institute: "MATH 204"
output:
  xaringan::moon_reader:
    css: [rladies, rladies-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(openintro)
library(faraway)
library(tidyverse)
library(ggformula)
library(ggmosaic)
library(patchwork)
library(latex2exp)
library(kableExtra)
library(broom)
```


# Inference for Categorical Data

In this lecture, we will examine two new inferential techniques:

--

- Testing for goodness of fit using chi-square, which is commonly used in two circumstances:

--

  - Given a sample of cases that can be classified into several groups, determine if the sample is representative of the general population. 
  
--

  - Evaluate whether data resemble a particular distribution, such as a normal distribution or a geometric distribution. 
  
--

- Testing for independence in two-way tables. 

---

# Learning Objectives




---

# Testing for goodness of fit: Motivating Example

```{r,echo=FALSE}
race_counts <- jury %>% group_by(race) %>% summarise(n=n())
jury_tb <- tibble(race=c("Representation in juries","Registered voters"),black=c(26,0.07),hispanic=c(25,0.12),white=c(205,0.72),other=c(19,0.09)) %>% mutate(total=black+hispanic+white+other)
jury_tb %>% kbl() %>% kable_styling()
```

We can compute the proportions for the data:

```{r,echo=FALSE}
race_counts %>% mutate(p = n/sum(n))
```